The Key to a Performant Data Parallel Compiler

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Add a discussion section

We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints). The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon .....

@@

>>>>

We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints). The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon  E5-1620v3.

@@

There are many options for representing graphs on data-parallel architecture. Most of them tend to optimize traversal instead of transformation. Because of the transformation heavy nature of compiler-style algorithms, these other representations can introduce work that complicates both the algorithms themselves, as well as the work necessary to maintain those structures over the lifetime of the compilation phase.

The use of a path matrix provides benefits to locality, which allows for increased parallelism for transformation passes such as the flattening passes mentioned above. While this can be used to increase the performance of a single compiler pass, the above treatment is not intended primarily for highly optimizing a single pass in a compiler. Instead, the design of the above algorithms and techniques are designed to improve the data parallelism and ease of expression across the entire compiler, not just a single pass.

To illustrate the difference, consider a few other possible representations. One possibility is the use of a single parent vector for the entire AST. We could also work directly off of the depth vector or the traditional method of listing pointers to the children and/or parents. In the case of the expression flattening above, we could likely achieve equal or better performance from using a single parent vector for this one computation. However, would be increasing costs elsewhere, particularly in maintenance. A path matrix provides both parent information, as well as information about the depth of the node, the siblings, position in the tree, and a unique identifier in a single structure that is cache/memory friendly. Since many operations on trees are memory bound, this memory optimization becomes important. In the Co-dfns compiler, we are able to generate the path matrix once, at the beginning of the compilation, and pass them around freely, and even duplicate them when appropriate, to communicate information without requiring cross thread or non-local memory access patterns.

This means that we never need to recompute the matrix, and indeed, doing so would defeat the purpose. In contrast, if you require a parent vector to organize your AST, then that parent vector may need to be shuffled, extended, and permuted each time the AST changes, and fails to maintain the original edge information of the graph. Keeping the original parent vector but maintaining a set of links would be a potential solution, but removes the locality of the access patterns, making it impossible to do a coalesced read of all of edge information needed for a single computation as a single access. Using a path matrix allows us to avoid that and increase the amount of stride-1 regular accesses and reduces memory indirections.

The cost of a path matrix is its size. Many other representations are size efficient, depending only on the number of nodes in the AST. A path matrix on the other hand, in the worst case, is dependent both on the depth of the AST and the number of nodes in the AST. Depending on the specific application, it is possible to choose different path matrix representations that optimize certain cases. For instance, one may not require that traversing the parent path be fast, and thus, we may reduce the cost of generating the path matrix by removing that requirement. Removing this requirement may also allow a path matrix to use a smaller sized integer representation for the matrix, further reducing the size at the cost of increased computation for certain operations. We have shown general path matrix techniques above that will work regardless of these particular choices in creating the path matrix.
